---
title: "EDX_Project"
author: "PP"
date: "2026-07-05"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 7)
```

## Introduction 

### Overview
The purpose of this project is to develop a movie recommendation system for individual users, akin to Netflix challenge and seen in Harvard Course 8 Machine Learning. 

Main method is based on machine learning based on the data made available by subset of Movielens, called edx. Key concepts shown in the course are explored which include movie recommendation system purely based on average rating and  5 features which are available through edx. The 5 features include "userId", "movieId", time data, "title" and "genres". Finally, this project will measure the final outcome in the form of residual mean squared error (RMSE). 

The assignment specifies the RMSE value of <0.86490 as full score worthy goal and this project shows how this could be achieved using 4 out of the 5 features. 

The English (Australian) style was used throughout this documentation given author's geographic background. 

### Basic data 
The basic standardised training and test data are provided. Code is provided in rmd file and not visualised below for ease of viewing.  
```{r echo=FALSE}
##########################################################
# Create edx and final_holdout_test sets 
##########################################################

# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(dplyr)
library(ggplot2)
library(dslabs)
library(lubridate)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

options(timeout = 120)

dl <- "ml-10M100K.zip"
if(!file.exists(dl))
  download.file("https://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings_file <- "ml-10M100K/ratings.dat"
if(!file.exists(ratings_file))
  unzip(dl, ratings_file)

movies_file <- "ml-10M100K/movies.dat"
if(!file.exists(movies_file))
  unzip(dl, movies_file)

ratings <- as.data.frame(str_split(read_lines(ratings_file), fixed("::"), simplify = TRUE),
                         stringsAsFactors = FALSE)
colnames(ratings) <- c("userId", "movieId", "rating", "timestamp")
ratings <- ratings %>%
  mutate(userId = as.integer(userId),
         movieId = as.integer(movieId),
         rating = as.numeric(rating),
         timestamp = as.integer(timestamp))

movies <- as.data.frame(str_split(read_lines(movies_file), fixed("::"), simplify = TRUE),
                        stringsAsFactors = FALSE)
colnames(movies) <- c("movieId", "title", "genres")
movies <- movies %>%
  mutate(movieId = as.integer(movieId))

movielens <- left_join(ratings, movies, by = "movieId")

# Final hold-out test set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.6 or later
# set.seed(1) # if using R 3.5 or earlier
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in final hold-out test set are also in edx set
final_holdout_test <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from final hold-out test set back into edx set
removed <- anti_join(temp, final_holdout_test)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)

```

## Method 

### Data cleaning 
Tidy data has been already provided by Edx team and this step was not required. 

### Data exploration and basic visualisations 
Basic understanding of the edx data is below: 
```{r}
#Basic information on the dataset 
str(edx)
dim(edx) #number of rows and columns 
edx %>% slice(1:10) #example of the first 10 rows
names(edx) #names of the heading 

n_distinct(edx$movieId) #distinct number of movies
n_distinct(edx$userId) #distinct number of users 

# The summary of ratings by each category 
edx %>% 
  select(rating) %>% 
  group_by(rating) %>% 
  summarise(frequencies = n())

# Sampling of 100 movies and 100 userId to check the extent of NA value - code is borrowed from Prof. 
users <- sample(unique(edx$userId), 100)
rafalib::mypar()
edx %>% filter(userId %in% users) %>% 
dplyr::select(userId, movieId, rating) %>%
mutate(rating = 1) %>%
pivot_wider(names_from = movieId, values_from = rating) %>% 
(\(mat) mat[, sample(ncol(mat), 100)])()%>%
as.matrix() %>% 
t() %>%
image(1:100, 1:100,. , xlab="Movies", ylab="Users")
abline(h=0:100+0.5, v=0:100+0.5, col = "grey")

# graphic representation of count of movies (graph_1) and those who rated (graph_2)
library(gridExtra)
graph_1 <- edx %>% count(movieId) %>% 
  ggplot(aes(n)) +
  geom_histogram(bins = 30, color = "black") +
  scale_x_log10() + 
  ggtitle("Movies")

graph_2 <- edx %>% count(userId) %>% 
  ggplot(aes(n)) +
  geom_histogram(bins = 30, color = "black") +
  scale_x_log10() + 
  ggtitle("Users")

grid.arrange(graph_1, graph_2, ncol = 2)

# numeric representation of count of movies (p1) and those who rated (p2)
edx %>% count(movieId) %>% summarise(mean = mean(n), min = min(n), Q1 = quantile(n, 0.25), median = median(n), Q3 = quantile(n, 0.75),  max= max(n))

edx %>% count(userId) %>% summarise(mean = mean(n), min = min(n), Q1 = quantile(n, 0.25), median = median(n), Q3 = quantile(n, 0.75),  max= max(n))

```

Above information tells us that
1. Data in the edx has 9000055 rows and 6 columns; 10677 distinct movies and 69878 raters. 
2. Rating is our dependent variable of interest. We have 5 other variables which may act as our predictors, and these could be explored to identify confounding/ correlation. 
3. Most of the ratings are whole numbers (e.g. 3, 4, 5) and that there is no 0. 
4. Similar to Movielens dataset, there are many NAs for which we need to create recommendation work
5. All movies had rating (including just 1 rating); all users rated >=10 times 

### RMSE and mathematical modeling approach  

#### RMSE 
RMSE is residual mean squared error. We define $y_{u,i}$ as the rating for movie $i$ by user $u$ and denote our prediction $\hat{y}_{u,i}$. 

The RMSE is then defined as following:
$RMSE = \sqrt{\frac{1}{n} \sum\limits_{u,i} (y_{u,i} - \hat{y}_{u,i})^2}$

RMSE can be interpreted similarly to a standard deviation. If this number is larger than 1, it means our typical error is larger than 1 star. 

#### Mathematical modeling approach

First model is based on simplest idea for the recommendation system. We can estimate this based on overall movie averages, which will be called $\mu$. Hence same rating for all movies and all users would be made. This mathematically looks like this: 

$Y_{u,i} = \mu + \epsilon_{u,i}$
where $\epsilon_{u,i}$ means errors sampled from the same distribution centered at 0. $\mu$ is the mean rating imputed for all movies.

Subsequent models will explore the impact of bias introduced by features mentioned above. 

$Y_{u,i} = \mu + b_i + b_u + b_t + b_g + \epsilon_{u,i}$

where 
1. $b_i$ is bias for the movie effects,
2. $b_u$, user effects, 
3. $b_t$, time effects, and
4. $b_g$, genre effects 

Further machine learning was planned including matrix factorisation, and principal component analysis; however this was deemed not necessary to achieve the desired outcome.

## Results 

### Model 1 - inputing average rating
Based on the above outlined mathematical modeling, first model looks at guessing the rating based on the average rating. 
Below tells us that RMSE of guessing the movie rating as an averagage of training set gives us RMSE of 1.061202.  
```{r}
# RMSE calculation function: 
RMSE <- function(rating_Y, ratings_P){
     sqrt(mean((rating_Y - ratings_P)^2)) #Y= true, P=predicted
}

#A first model
mu <- mean(edx$rating) #this is mu_hat
mu #3.512465

model_1_rmse <- RMSE(final_holdout_test$rating, mu)
model_1_rmse

rmse_results <- tibble(Method = "Model 1. Average", RMSE = model_1_rmse) 

knitr::kable(rmse_results)
```

### Model 2 - individual movie impact 

Some movies are generally rated higher than others. We can supplement our previous model by adding the term $b_i$ to represent average ranking for movie $i$. The combined model improves the RMSE to 0.9439087.

```{r}
# mu is from earlier chunk
movie_mean <- edx %>% 
     group_by(movieId) %>% 
     summarize(b_i = mean(rating - mu))

#graph representation of the b_i (with 1.5 being the highest possible value given mean of 3.5, and -3.0 being the minimal given the lowest rating of 0.5)
edx %>% 
     group_by(movieId) %>% 
     summarize(b_i = mean(rating - mu)) %>% 
  ggplot(aes(b_i)) +
  geom_histogram(bins = 30, color = "black")

# additional impact of b_i on RMSE 
rating_P <- mu + final_holdout_test %>% 
     left_join(movie_mean, by='movieId') %>%
     pull(b_i)

model_2_rmse <- RMSE(rating_P, final_holdout_test$rating)
rmse_results <- bind_rows(rmse_results,
                          tibble(Method="Model 2. Movie Effect Model",
                                     RMSE = model_2_rmse ))
knitr::kable(rmse_results) 

```

### Model 3 - individual rater impact 

Some raters rate movies higher than others. We can supplement our previous model by adding the term $b_u$ to represent average ranking for users $u$. The combined model improves the RMSE to 0.8653488, which is a significant and massive improvement. 
```{r}
#individual rater average rating by counts
edx %>% 
     group_by(userId) %>% 
     summarize(b_u = mean(rating)) %>%
     ggplot(aes(b_u)) + 
     geom_histogram(bins = 30, color = "black")

# additional impact of b_u on RMSE 
user_mean <- edx %>% 
     left_join(movie_mean, by='movieId') %>%
     group_by(userId) %>%
     summarise(b_u = mean(rating - mu - b_i))

rating_P <- final_holdout_test %>% 
     left_join(movie_mean, by='movieId') %>%
     left_join(user_mean, by='userId') %>%
     mutate(predicted = mu + b_i + b_u) %>%
     pull(predicted)

model_3_rmse <- RMSE(rating_P, final_holdout_test$rating)

rmse_results <- bind_rows(rmse_results,
                          tibble(Method="Model 3. Movie + User Effects Model",  
                                     RMSE = model_3_rmse ))

knitr::kable(rmse_results) 
```

Due to a major drop in RMSE with $b_u$ addition, it is worthwhile checking for any potential errors.  

```{r}

# distribution of user effect appear reasonable 
summary(user_mean$b_u) 

# number of ratings were shown previously and reproduced below. There are no raters with 1-3 rated movies; Most have rated more than 10 movies. 
edx %>% count(userId) %>% summarise(mean = mean(n), min = min(n), Q1 = quantile(n, 0.25), median = median(n), Q3 = quantile(n, 0.75),  max= max(n))

edx %>% count(userId) %>% ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") +
  scale_x_log10() 

# Below is testing for over-fitting by testing for RMSE on the train set. The result does not support major difference between RMSE of the test and train set and therefore minimal over-fitting. 

predicted_train_ratings <- edx %>% 
     left_join(movie_mean, by='movieId') %>%
     left_join(user_mean, by='userId') %>%
     mutate(predicted = mu + b_i + b_u) %>%
     pull(predicted)

train_rmse <- RMSE(predicted_train_ratings, edx$rating)

train_rmse
```


### Impact of time/date 

Timing of movie release would impact on the rating. The relationship with time to movies are explored below, which shows that (1) relationship between time and rating is complicated. Overall rating  decreased over the time of movie releases. In other words, more recent the movies, lower the rating. But (2) there was plateau of the rating decline by 2000, and there is up-trend in terms of rating since 2005. 

```{r}
#create date column separated by "week"
edx <- edx %>% 
  mutate(date = as_datetime(timestamp)) %>% 
  mutate(date = floor_date(date, unit = "week")) #floor_date better than round_date for week specific effects 

#similar operation is needed for the test_set 
final_holdout_test <- final_holdout_test %>% 
  mutate(date = as_datetime(timestamp)) %>% 
  mutate(date = floor_date(date, unit = "week"))

#visualisation
edx %>% mutate(date = floor_date(date, unit = "week")) %>%
	group_by(date) %>%
	summarize(rating = mean(rating)) %>%
	ggplot(aes(date, rating)) +
	geom_point() +
	geom_smooth()
```
We can further supplement our previous model by adding the term $b_t$ to represent average time impact for time $t$ measured in weeks so that we can remove noise of individual date of movie release. The combined model improves the RMSE to 0.8652433	which is only a slight improvement. 

```{r}
#building the prediction model 

date_mean <- edx %>% 
     left_join(movie_mean, by='movieId') %>%
     left_join(user_mean, by='userId') %>%
     group_by(date) %>%
     summarise(b_t = mean(rating - mu - b_i - b_u))

rating_P <- final_holdout_test %>% 
     left_join(movie_mean, by='movieId') %>%
     left_join(user_mean, by='userId') %>%
    left_join(date_mean, by='date') %>%
     mutate(predicted = mu + b_i + b_u + b_t) %>%
     pull(predicted)

model_4_rmse <- RMSE(rating_P, final_holdout_test$rating)

rmse_results <- bind_rows(rmse_results,
                          tibble(Method="Model 4. Movie + User + Date Effects Model",  
                                     RMSE = model_4_rmse ))

knitr::kable(rmse_results) 

```

### Impact of genres

Genres are provided in the edx data set. Some genres are liked by users more than the others. This relationship with genres to movies are explored below. 

```{r}
#there is a clear effect of genres - please excuse the obscuration of the x-axis labels - this will be more clearly demonstrated in the next section. 
edx %>%
    group_by(genres) %>%
    summarize(mean_rating = mean(rating), n = n(), se = sd(rating)/sqrt(n), title=title[1]) %>% 
  filter(n>=1000)  %>% 
  mutate(genres = reorder(genres, mean_rating)) %>% #arrange(mean_rating) this does not work
  ggplot(aes(x = genres, y = mean_rating, ymin = mean_rating - 2*se, ymax = mean_rating + 2*se)) + 
	geom_point() +
	geom_errorbar() + 
	theme(axis.text.x = element_text(angle = 90, hjust = 1))

#10 lowest and maxmimum rated genres with ratings >1000:

edx %>% group_by(genres) %>%
	summarize(mean_rating = mean(rating), n = n(), se = sd(rating)/sqrt(n), title=title[1]) %>% 
	filter(n >= 1000) %>% 
	mutate(genres = reorder(genres, mean_rating)) %>% 
  slice_min(n=10, genres)

edx %>% group_by(genres) %>%
	summarize(mean_rating = mean(rating), n = n(), se = sd(rating)/sqrt(n), title=title[1]) %>% 
	filter(n >= 1000) %>% 
	mutate(genres = reorder(genres, mean_rating)) %>% 
  slice_max(n=10, genres)
```

We can further supplement our previous model by adding the term $b_g$ to represent average genre impact for genre $g$ measured. Rather than teasing out individual genre impacts (i.e. if there is combination of comedy and action, disecting these out individually), genres as they are have been kept. The combined model improves the RMSE to 0.8648418	which is only a slight improvement but which achieves the RMSE that gets us the full mark. 

```{r}
#code for the model 

genre_mean <- edx %>% 
     left_join(movie_mean, by='movieId') %>%
     left_join(user_mean, by='userId') %>%
     left_join(date_mean, by='date') %>%
     group_by(genres) %>%
     summarise(b_g = mean(rating - mu - b_i - b_u - b_t))

rating_P <- final_holdout_test %>% 
     left_join(movie_mean, by='movieId') %>%
     left_join(user_mean, by='userId') %>%
    left_join(date_mean, by='date') %>%
    left_join(genre_mean, by='genres') %>%
     mutate(predicted = mu + b_i + b_u + b_t + b_g) %>%
     pull(predicted)

model_5_rmse <- RMSE(rating_P, final_holdout_test$rating)

rmse_results <- bind_rows(rmse_results,
                          tibble(Method="Model 5. Movie + User + Date + Genre Effects Model",  
                                     RMSE = model_5_rmse ))

knitr::kable(rmse_results) 
```

## Conclusion 
This project used RMSE as a measure of the loss function to see the impact of 4 out of 5 available features in edx data set. Using machine learning techniques, we were able to build movie recommendation system which can achieve RMSE of 0.8648418 stars. Although not perfect, the margin of error is within 1 star. 

One major limitation is that more advanced machine learning techniques were not used in this instance. Partly this was due to limited time avaibility and the fact that Netflix team's winning RMSE of 0.8660 following extensive and advanced analysis using the Netflix data was slightly higher than the RMSE achieved with basic analysis performed in this project. Future projects could look at further use of advanced analysis on edx dataset. 

In summary, use of basic machine learning methods achieved RMSE of 0.8648418 stars for edx dataset. 

## References 

1. The backbone of the analysis was based on Prof Rafael A. Irizarry, available in the text book chapter 33 at https://rafalab.dfci.harvard.edu/dsbook/

2. Netflix challenge paper with a more detailed explanation was reviewed and available here: https://www2.seas.gwu.edu/~simhaweb/champalg/cf/papers/KorenBellKor2009.pdf

3. Large language model, ChatGPT, was used for checking for coding errors. 